Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-31 pnas/Towards.tex
Update Appendix D
Update appendices

\TODO{Need to cite \cite{scapin2018cryo}. Discusses cryo-em for drug discovery and mentions small molecule}
%

@@

We assume the binary signal $s$ obeys a \emph{separation condition}:
%% Nicolas: this is too important: it must stand out
% If $s[i] = 1$ and $s[j] = 1$  for  $i \neq j$, then $\vert i - j\vert \geq 2L-1$.
\begin{align} 
\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

Our problem can be interpreted as a special case of the \emph{system identification} problem. Similarly to~\eqref{eq:model}, the
%system identification
forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the system's impulse response), $w$ is an unknown (often random) input sequence, and $\varepsilon$ is an additive noise.
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w$ is binary and satisfies our separation condition, we recover our model. 
%The same observation model is used for \emph{blind deconvolution}, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few.

Likelihood-based methods estimate $x$ as the maximizer of some function $f(x | y)$, where $f$ is derived from the likelihood function of $x$ given the observed signal $y$. %For example, $f$ may be the likelihood itself, or a related function with a similar form (leading to the class of ``quasi-likelihood'' methods). 
If some prior is assumed on $x$, then $f(x|y)$ can be taken to be the posterior distribution of $x$ given the data; this is the simplest form of Bayesian inference.
Methods based on such formulations are popular nowadays in cryo-EM; see for instance~\cite{sigworth1998maximum,scheres2012relion}. 
Optimizing the function $f(x|y)$ exactly is often intractable, and thus heuristic methods are used instead. One proposed technique is to use Markov Chain Monte Carlo (MCMC)~\cite{cappe1999simulation}. %Another paper considers parameterized models for multiple distinct signals, as in our framework ($K>1$)~\cite{andrieu2001bayesian}. Their proposed solution is an MCMC algorithm tailored for their specific parametrized problem. 
In special cases, including the case where $w$ is binary, expectation maximization (EM) has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based on a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is superlinear in $N$, and therefore its usage is limited for big data sets. 
%Indeed, for a general support signal $s$, on each iteration of EM, a probability must be assigned to any feasible combination of positions for the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$---$O(N^M)$ such combinations in total. Hence, the problem becomes computationally intractable when $M$ grows with $N$ and $N$ is large. However, exploiting the separation condition on the support~\eqref{eq:spacing}, we devise in Section TKTK an approximate EM procedure for  our model. 
%
%Because likelihood methods are computationally expensive, methods based on recovery from moments, which are akin to our method, have also been previously used for system identification. Methods based on the third- and fourth-order moments are described and analyzed in~\cite{lii1982deconvolution,giannakis1989identification,tugnait1984identification}.

@@

To support this claim, we consider a strictly simpler problem: suppose an oracle identifies for us one interval of length $L$ in the micrograph that either contains a full signal occurrence (plus noise), or contains just noise. Our task is to determine which one it is, that is, to determine whether the corresponding entry of $s$ is 0 or 1.
The oracle further provides the signal $x$, the probability $q$ that  the interval contains signal, and the noise variance $\sigma^2$. %\TODO{Nico: I don't see how the next sentence is relevant for the task of estimating $s$---Since we have the full generative model, we can always generate more data as needed; therefore it is enough to consider the hardness of deciding whether a single interval contains signal or not.---I would just write this: Since the intervals are independent of each other, it is sufficient to consider a single interval.}

This decision problem can be abstracted as follows: We have two known vectors $\theta_0 = x$ and $\theta_1 = 0$ in $\mathbb{R}^L$. There is a random variable $\eta$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \mathbb{R}^L$ (an extract of the micrograph) with the following distribution:
% $X | (\eta  = i) \sim N(\theta_i,\sigma^2)$; in other words,
if $\eta = 0$, then $X \sim \mathcal N(\theta_0,\sigma^2)$ %(that is, $X$ follows a normal distribution around $\theta_0$ with variance $\sigma^2$),
 and if $\eta = 1$, then $X \sim \mathcal N(\theta_1,\sigma^2)$. %\TODO{Tamir: to remove parenthesis}
%If $\theta_1=0$, then the latter corresponds to pure noise interval.

@@

Proposition~\ref{prop:two_gauss} implies that in order to estimate the signal at low SNR we must consider methods that aim to estimate the signal $x$ directly, without estimating the nuisance variable $s$ as an intermediate step.
%we need to consider a different approach.
In the next sections, % Nicolas: sections is plural on purpose
we consider autocorrelation analysis for that purpose.
%In what follows, we present several results on properties of autocorrelations.

@@

For shifts in $0, \ldots, L-1$, the autocorrelation functions of $y$ depend on the corresponding autocorrelations of $x$, the noise level $\sigma$ and the support signal $s$. Importantly, under the separation condition~\eqref{eq:spacing}, the dependency on $s$ is only through the density $\gamma$.
%In particular, for the first-order autocorrelation (that is, the mean), since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:
%\begin{align*}
%\E\{ a_y^1 \} & = \gamma a_x^1.
%\end{align*}
%(The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic: under the conditions, the quantity under consideration depends on $s$ only through $\gamma$.)
We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density). In that regime, the law of large numbers can be used to show the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & \aseq \gamma a_{x}^1,
\end{align}
where equality holds almost surely (a.s.), meaning it holds with probability one.
The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic.
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable: if $z[0]$ or $z[L-1]$ are close to 0, recovery of $z$ is sensitive to errors in the autocorrelations. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the separation condition holds, then the length of the signal can be determined from the autocorrelations.
In particular, if  $a_y^2[L-i] = 0$ for $i=1,\ldots,T$,
then the support of $x$ has length $L-T$ rather than $L$. 
%the separation condition holds 
%\TODO{for some spacing $W\geq L$---Nico: what is this?}, then $a_z^2[i] = 0$ for all $i > L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (or phase for complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we can directly deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	 The signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are identifiable from the first three autocorrelation functions of $y$ if:
	\begin{itemize}
		\item Either the signal $x$ is generic; or
		\item   $x$ has nonzero mean  and $\gamma\geq\frac{L}{4(L-1)}$.
	\end{itemize}
\end{corollary}

%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

In the first experiment, we estimated 
an 50-by-50 pixel image of Einstein with mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. Each micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$,  corresponding  to SNR = $\frac{M\|x\|_F^2} {\sigma^2N} \approx1/370$. This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with the separation condition~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

@@

%\begin{figure*}[h!]
%	\centering
%	\includegraphics[scale=0.7]{Einstein_progress_examples}
%	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
%		%\TODO{To add: redo the figures according to Amit's comments and the figures of noisy Einstein's micrographs}
%		% At a noise level of $\sigma = 3$, this amounts to an SNR of $1/20$.
%	}
%\end{figure*}

@@

%\begin{figure}[h]
%	\centering
%\begin{subfigure}[h]{0.25\textwidth}
%	\centering
%	\includegraphics[scale=0.33]{Einstein_ps_error}
%	\caption{relative power spectrum estimation error}
%\end{subfigure} 
%\begin{subfigure}[h]{0.25\textwidth}
%	\centering
%	\includegraphics[scale=0.33]{Einstein_recovery_error}
%	\caption{relative recovery error}
%\end{subfigure} 
%	\caption{\label{fig:error_per_micro} Error curves for the experiment of Figure~\ref{fig:Einst_example}}
%\end{figure}

%
%\begin{figure}[h]
%	\centering
%	\includegraphics[scale=.55]{Einstein_progress} (left) power spectrum estimation error (right) RRR error after fixed number of 1000 iterations.
%	\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
%\end{figure}

@@

While we cannot provide a 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics (e.g.,~\cite{donoho2004higher}) and debates in the cryo-EM community~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.
%
%Under the experimental parameters (see Section~\ref{sec:numeric_details}), such values of $\gamma$ predict  in the informative micrograph and less than one projection in the pure noise scenario.

@@

%In the simplified model for cryo-EM we examined, we showed it is possible to estimate the 3-D structure of small volumes. 
In the simplified mathematical model we examined, we showed it is possible to estimate a signal without detecting its appearances.  
Our strategy is to compute autocorrelation functions of the micrographs and to relate these statistics to the unknown signal's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.

@@

The numerical experiments we have performed reveal that the third-order autocorrelation may not be enough for 3-D reconstruction in practice, due to high sensitivity.
This suggests that fourth-order autocorrelation may be necessary. This in turn would imply that the procedure might require a large amount of data.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable SNR remains an open question.} Recent trends in high-throughput cryo-EM technology   give hope that this may be a lesser concern in the long term. Still, large amounts of data also require large amounts of computation. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one pass through each micrograph is necessary. The output of this data processing stage is a summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. %The proof-of-concept experiments above suggest they might. \TODO{Is it?}

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generated micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we placed Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition.%~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation condition and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an SNR of approximately $1/370$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

@@

To recover the target image from the estimated power spectrum, we use a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z)),
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, possibly reflected through its origin, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner. In all experiments, the algorithm halted after a fixed number of 2000 iterations.
%
%In order to compare the performance in multiple cases and at different noise levels,  and the iterate with the smallest error compared to the ground truth (up to the reflection ambiguity) is chosen as the solution. While this cannot be done in practice (since we do not have access to the ground truth to determine which iterate is best), this procedure enables us to compare a large number of instances in different noise environments. \TODO{Note the last two sentences!}

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.
%	 fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
%This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$
%% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/9.

%This is enough noise to make cross-correlations of $y$ even with the true signals display peaks at essentially random locations, uninformative of the actual locations of the signal occurrences. Thus, we contend that it would be difficult for any algorithm to locate the signal occurrences, let alone to classify them according to which signal appears where.

%\TODO{TB: I would place the equation counting argument in the theory section in the paragraph of open questions (I marked th place)}
%
Given the observation $y$, we proceed to compute the autocorrelations. %The first-order autocorrelation is straightforward. 
We excluded biased elements of the autocorrelations; that is $a_y^2[\ell_1]$ and $a_y^3[\ell_1, \ell_2]$ such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero. 
This has the  practical effect that we need not know $\sigma$. 
%For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---
By also taking symmetries into account, the third-order autocorrelation contains $\frac{(L-1)(L-2)}{2}$ remaining entries. Thus, in total we have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients. 
This redundancy suggests that it might be possible to estimate several signals simultaneously (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}).
%Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
%\begin{align*}
%K(L+1) \leq \frac{1}{2} L (L-1) + 1.
%\end{align*}
%Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$ (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}). 
In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ (if the length of the measurement is larger than that) and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

>>>>

\TODO{Need to cite \cite{scapin2018cryo}. Discusses cryo-em for drug discovery and mentions small molecule}

@@

We assume the binary signal $s$ obeys a \emph{separation condition}:
\begin{align} 
\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

Our problem can be interpreted as a special case of the \emph{system identification} problem. Similarly to~\eqref{eq:model}, the
forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the system's impulse response), $w$ is an unknown (often random) input sequence, and $\varepsilon$ is an additive noise.
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w$ is binary and satisfies our separation condition, we recover our model.

Likelihood-based methods estimate $x$ as the maximizer of some function $f(x | y)$, where $f$ is derived from the likelihood function of $x$ given the observed signal $y$.  
If some prior is assumed on $x$, then $f(x|y)$ can be taken to be the posterior distribution of $x$ given the data; this is the simplest form of Bayesian inference.
Methods based on such formulations are popular nowadays in cryo-EM; see for instance~\cite{sigworth1998maximum,scheres2012relion}. 
Optimizing the function $f(x|y)$ exactly is often intractable, and thus heuristic methods are used instead. One proposed technique is to use Markov Chain Monte Carlo (MCMC)~\cite{cappe1999simulation}. 
In special cases, including the case where $w$ is binary, expectation maximization (EM) has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based on a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is superlinear in $N$, and therefore its usage is limited for big data sets.

@@

To support this claim, we consider a strictly simpler problem: suppose an oracle identifies for us one interval of length $L$ in the micrograph that either contains a full signal occurrence (plus noise), or contains just noise. Our task is to determine which one it is, that is, to determine whether the corresponding entry of $s$ is 0 or 1.
The oracle further provides the signal $x$, the probability $q$ that  the interval contains signal, and the noise variance $\sigma^2$.

This decision problem can be abstracted as follows: We have two known vectors $\theta_0 = x$ and $\theta_1 = 0$ in $\mathbb{R}^L$. There is a random variable $\eta$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \mathbb{R}^L$ (an extract of the micrograph) with the following distribution: if $\eta = 0$, then $X \sim \mathcal N(\theta_0,\sigma^2)$ and if $\eta = 1$, then $X \sim \mathcal N(\theta_1,\sigma^2)$.

@@

Proposition~\ref{prop:two_gauss} implies that in order to estimate the signal at low SNR we must consider methods that aim to estimate the signal $x$ directly, without estimating the nuisance variable $s$ as an intermediate step.
In the next sections, we consider autocorrelation analysis for that purpose.

@@

For shifts in $0, \ldots, L-1$, the autocorrelation functions of $y$ depend on the corresponding autocorrelations of $x$, the noise level $\sigma$ and the support signal $s$. Importantly, under the separation condition~\eqref{eq:spacing}, the dependency on $s$ is only through the density $\gamma$.
We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density). In that regime, the law of large numbers can be used to show the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & \aseq \gamma a_{x}^1,
\end{align}
where equality holds almost surely (a.s.), meaning it holds with probability one.
The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic.
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable: if $z[0]$ or $z[L-1]$ are close to 0, recovery of $z$ is sensitive to errors in the autocorrelations. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the separation condition holds, then the length of the signal can be determined from the autocorrelations.
In particular, if  $a_y^2[L-i] = 0$ for $i=1,\ldots,T$,
then the support of $x$ has length $L-T$ rather than $L$. 
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (or phase for complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we can directly deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	 The signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are identifiable from the first three autocorrelation functions of $y$ if:
	\begin{itemize}
		\item Either the signal $x$ is generic; or
		\item   $x$ has nonzero mean  and $\gamma\geq\frac{L}{4(L-1)}$.
	\end{itemize}
\end{corollary}
%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

In the first experiment, we estimated 
an 50-by-50 pixel image of Einstein with mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. Each micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$,  corresponding  to SNR = $\frac{M\|x\|_F^2} {\sigma^2N} \approx1/370$. This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with the separation condition~\eqref{eq:spacing}.

@@

@@

@@

While we cannot provide a 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics (e.g.,~\cite{donoho2004higher}) and debates in the cryo-EM community~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.

@@

In the simplified mathematical model we examined, we showed it is possible to estimate a signal without detecting its appearances.  
Our strategy is to compute autocorrelation functions of the micrographs and to relate these statistics to the unknown signal's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.

@@

The numerical experiments we have performed reveal that the third-order autocorrelation may not be enough for 3-D reconstruction in practice, due to high sensitivity.
This suggests that fourth-order autocorrelation may be necessary. This in turn would imply that the procedure might require a large amount of data.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable SNR remains an open question.} Recent trends in high-throughput cryo-EM technology   give hope that this may be a lesser concern in the long term. Still, large amounts of data also require large amounts of computation. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one pass through each micrograph is necessary. The output of this data processing stage is a summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice.

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generated micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we placed Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition ~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation condition and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an SNR of approximately $1/370$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.

@@

To recover the target image from the estimated power spectrum, we use a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z)),
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, possibly reflected through its origin, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner. In all experiments, the algorithm halted after a fixed number of 2000 iterations.

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.

Given the observation $y$, we proceed to compute the autocorrelations. %The first-order autocorrelation is straightforward. 
We excluded biased elements of the autocorrelations; that is $a_y^2[\ell_1]$ and $a_y^3[\ell_1, \ell_2]$ such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero. 
This has the  practical effect that we need not know $\sigma$. 
By also taking symmetries into account, the third-order autocorrelation contains $\frac{(L-1)(L-2)}{2}$ remaining entries. Thus, in total we have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients. 
This redundancy suggests that it might be possible to estimate several signals simultaneously (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}).
In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ (if the length of the measurement is larger than that) and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

Having computed the moments of interest, we now estimate the signal $x$ and coefficient $\gamma$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear LS problem:
%\begin{align}
%\min_{\substack{\hat x_1, \ldots, \hat x_K \in \mathbb{R}^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}}& w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 \\ & + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2  \nonumber \\&+ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2, \nonumber
%\label{eq:optim1D}
%\end{align}
\begin{align}
\min_{\substack{\hat x\in \mathbb{R}^{W} \\ \hat \gamma > 0}}& w_1 \left( a_y^1 - \hat \gamma a_{\hat x}^1 \right)^2  + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \hat \gamma a_{\hat x}^2[\ell] \right)^2  \nonumber \\&+ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] -  \hat \gamma a_{\hat x}^3[\ell_1,\ell_2] \right)^2, 
\label{eq:optim1D}
\end{align}
where $W \geq L$ is the length of the sought signal and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).
